---
title: "NFL Play by Play"
output: html_notebook
---

Load packages

```{r}
library(readxl) 
library(writexl)
library(dplyr)
library(tree)
library(boot)
library(caret)
library(RANN)
library(e1071)
library(arm)
library(xgboost)
library(lattice)
library(rattle)
library(kernlab)
library(rpart)
library(ggcorrplot)
library(corrplot)
library(MASS)
library(pROC)
library(mice)
library(fastDummies)
```

Load datasets 

```{r}
pbp_13 = read_excel("nfl_pbp_13.xlsx")
pbp_14 = read_excel("nfl_pbp_14.xlsx")
pbp_15 = read_excel("nfl_pbp_15.xlsx")
pbp_16 = read_excel("nfl_pbp_16.xlsx")
pbp_17 = read_excel("nfl_pbp_17.xlsx")
pbp_18 = read_excel("nfl_pbp_18.xlsx")
pbp_19 = read_excel("nfl_pbp_19.xlsx")
```

Data Cleaning and Manipulation

```{r}
pbp_13 = pbp_13 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_13$GameDate = as.Date(pbp_13$GameDate)
pbp_13 = pbp_13[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_14 = pbp_14 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_14$GameDate = as.Date(pbp_14$GameDate)
pbp_14 = pbp_14[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_15 = pbp_15 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_15$GameDate = as.Date(pbp_15$GameDate)
pbp_15 = pbp_15[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_16 = pbp_16 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_16$GameDate = as.Date(pbp_16$GameDate)
pbp_16 = pbp_16[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_17 = pbp_17 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_17$GameDate = as.Date(pbp_17$GameDate)
pbp_17 = pbp_17[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_18 = pbp_18 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_18$GameDate = as.Date(pbp_18$GameDate)
pbp_18 = pbp_18[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]

pbp_19 = pbp_19 %>% filter(IsTwoPointConversion == "0") %>%
  filter(IsPenalty == "0") %>%
  filter(IsChallenge == "0") %>%
  filter(IsSack == "0") %>%
  filter(IsIncomplete == "0") %>%
  filter(IsInterception == "0") %>%
  filter(IsFumble == "0") %>%
  filter(Formation %in% c("NO HUDDLE", "NO HUDDLE SHOTGUN", "SHOTGUN", "UNDER CENTER", "WILDCAT")) %>%
  filter(PlayType %in% c("RUSH", "PASS", "SCRAMBLE"))
pbp_19$GameDate = as.Date(pbp_19$GameDate)
pbp_19 = pbp_19[c(1,2,3,4,6,7,8,9,19,20,21,23,24,27,38,39,40,12)]
```

Combine datasets to make one large train set - 2013 to 2018

```{r}
pbp_combined = pbp_13 %>% rbind(pbp_14) %>%
  rbind(pbp_15) %>%
  rbind(pbp_16) %>%
  rbind(pbp_17) %>%
  rbind(pbp_18)
```

Fill in NAs for train set and select columns to be used in the train set

```{r}
pbp_combined$PassType[is.na(pbp_combined$PassType)] = 'OTHER'
pbp_combined$RushDirection[is.na(pbp_combined$RushDirection)] = "OTHER"

pbp_train = pbp_combined[c(3,4,7,8,11,13,14,15,16,17,18)]
```

Fill in NAs for test set and select columns to be used in the test set

```{r}
# FIll in NAs for test set
pbp_19$PassType[is.na(pbp_19$PassType)] = 'OTHER'
pbp_19$RushDirection[is.na(pbp_19$RushDirection)] = "OTHER"

# Select columns to be used in test set
pbp_test = pbp_19[c(3,4,7,8,11,13,14,15,16,17,18)]
```

Remove character variables for correlation plot

```{r}
pbp_cor = pbp_train[-c(5,7,8,10)]

corrplot(cor(pbp_cor, use="complete.obs"), method="number",type="lower", number.cex = .7)
```

Change character variables to factors and fix column names for modeling

```{r}
pbp_train$Formation = as.factor(pbp_train$Formation)
pbp_train$PassType = as.factor(pbp_train$PassType)
pbp_train$RushDirection = as.factor(pbp_train$RushDirection)
pbp_train$YardLineDirection = as.factor(pbp_train$YardLineDirection)
pbp_train$SeriesFirstDown = as.factor(pbp_train$SeriesFirstDown)

pbp_test$Formation = as.factor(pbp_test$Formation)
pbp_test$PassType = as.factor(pbp_test$PassType)
pbp_test$RushDirection = as.factor(pbp_test$RushDirection)
pbp_test$YardLineDirection = as.factor(pbp_test$YardLineDirection)
pbp_test$SeriesFirstDown = as.factor(pbp_test$SeriesFirstDown)

colnames(pbp_train) <- make.names(colnames(pbp_train))
colnames(pbp_test) <- make.names(colnames(pbp_test))
```

Logistic regression analysis for variable significance

```{r}
pbp.logi.1 = glm(SeriesFirstDown~., pbp_train, family="binomial")
summary(pbp.logi.1)
```

ML models with 10 fold cross validation

```{r}
# 10-fold Cross-Validation
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"

# Linear Discriminant Analysis (LDA)
set.seed(1)
fit.lda.2 <- train(SeriesFirstDown~., data=pbp_train, method="lda", metric=metric, trControl=control)

# Classfication and Regression Trees (CART)
set.seed(1)
fit.cart.2 <- train(SeriesFirstDown~., data=pbp_train, method="rpart", metric=metric, trControl=control)

# Bayesian Generalized Linear Model - Logistic Regression
set.seed(1)
fit.baylogi.2 <- train(SeriesFirstDown~., data=pbp_train, method="bayesglm", metric=metric, trControl=control)

# Logistic Regression
set.seed(1)
fit.logi.2  <- train(SeriesFirstDown~., data=pbp_train, method="glm", metric=metric, trControl=control)

```

Summarize accuracy of models and select the best model

```{r}
results.1 <- resamples(list(lda=fit.lda.2, cart=fit.cart.2, baylogi=fit.baylogi.2, logi=fit.logi.2))
summary(results.1)
```

Summarize the Best Model

```{r}
print(fit.baylogi.2)
```

Accuracy of the model predicting the dependent variable in the test set

```{r}
pbp_test_1 = pbp_test[-c(11)]

pbp_test_1$SeriesFirstDown1 = predict(fit.baylogi.2, type = "raw", newdata=pbp_test_1)
mean(pbp_test_1$SeriesFirstDown1==pbp_test$SeriesFirstDown)
table(pbp_test_1$SeriesFirstDown1,pbp_test$SeriesFirstDown)
```

ROC Curve, Threshold, Sensitivity, and Specificity

```{r}
pred.baylogi = predict(fit.baylogi.2, type = "prob", pbp_train)
pred.baylogi_1 = as.numeric(pred.baylogi[,2])
baylogi.roc = roc(response = pbp_train$SeriesFirstDown, predictor = pred.baylogi_1)
plot(baylogi.roc, legacy.axes = TRUE, print.auc.y = 1.0, print.auc = TRUE)
coords(baylogi.roc, "best", "threshold")
```

